---
title: "Diabetes hospital admission"
author: "Neha"
date: "25/11/2021"
output:
  pdf_document: default
always_allow_html: true
  #html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The objective of reproducibility in research can be achieved by combining specific instruction to experimental data and data analysis in one place for better understanding, verification, and recreation. This can be achieved by using R markdown which allows use of document that is combination of content and data analysis code. The Sweave function and the knitr package can be used to blend the subject matter and R code in a single document that defines content and the analysis. [3]


R markdown is a file format within R which allows creation of reproducible and dynamic documents It consist of three types of content YAML header,text and code chunk .R markdown can produce output/report in word, HTML, pdf, dashboards and slideshows etc which can be easily shared.The output is defined in the YAML header.


YAML header is a short text which gives information of title, author, date and output demarcated by three dashes (—) on either end of the markdown file 


R markdown provides syntax to create different levels of headers. For example, the word “introduction” of this report has been created as main heading by using single “#” before the text “introduction” and other levels of headings can be created by increasing the number of “#” in front of the text. For example 



superscripts ^2^ by using “^” 


~~strikethrough~~ by using “~~” 



Unordered lists ca be written as follows by using “*” in front of the text and “+” in front of sub-unordered lists



* Unordered list 
* Item 2
    + sub-item 1
    + sub-item 2

Ordered list can be written as following by writing order like “1.” and sub items using “+” in front of the text

1. ordered list 
2. item 2 
      + sub-item 1
      + sub-item 2

Tables can also be created by using "-" to underline the table header and then use  "|" to separate table content on each side.

 Table Header | Second Header
------------- | ------------- 
       Cell 1 | Cell 2 
       Cell 3 | Cell 4


To create a blockquote, add a “> “in front of a paragraph. For example 

   >  This is blockquote

R markdown facilitate code chunk which is simply a display of code for illustration and its output. For example, below code chunk is for loading the different packages used in this report.The code chunk is used for detailed snippets of code and it is achieved by using three back ticks in start of the code and finish with 3 back ticks 


Chunk output can be customised with arguments set within the curly bracket { } for example 


include = FALSE prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.


echo = FALSE prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.


message = FALSE prevents messages that are generated by code from appearing in the finished file.


warning = FALSE prevents warnings that are generated by code from appearing in the finished.


R markdown also facilitates inline code by putting back ticks (') around part of the code line. In line code is usually for small command within the test of the report.

  

```{r load_package,  message= FALSE}
library(tidyverse)
library(plotly)    # interactive visualisations
library(eeptools)  # calculating age from DOB
library(viridis)   # nice color scheme
library(here)
```


. A comma separated values data can be read though read_csv function 




```{r read-data , message = FALSE}


demographic <- read_csv(here("data/demographics.csv"))
Hospital_admission <- read_csv(here("data/Diabetes related hospital admission.csv"))

```



```{r missing-values}
glimpse(demographic)
glimpse(Hospital_admission)

```

checking missing values

```{r missing-value-demographic}

demographic %>% summary()
demographic$sex %>%table()   # any missing value in sex

```


```{r missing-values-hospital-admission}

Hospital_admission$result %>%table()  # any missing value in results

```


```{r tpos-in-data}

Hospital_admission %>% summary()  # any typos in hospital admission data set

```




```{r remove-missing-values}

tidy_hospital_admission <- Hospital_admission %>%
  drop_na()%>% 
  filter(result != '*')

tidy_hospital_admission$result %>% table()

tidy_hospital_admission <- Hospital_admission %>% 
  drop_na()%>% 
  filter(value != '10')

tidy_hospital_admission %>% summary()
```



joining the two data set

```{r join-two-dataset}

admission_data <- left_join(tidy_hospital_admission,demographic,by = c("chi" = "chi-number")) 

admission_data <- admission_data %>% 
  rename(Sugar_reading = value)

glimpse(admission_data)

```


save the clean data 

```{r  save-clean-data}

write_csv(admission_data, "Sugar reading.csv")

```




```{r histogram}

#histogram

admission_data %>% 
    ggplot(aes(x= Sugar_reading))+
  geom_histogram( bins = 15,color = "black", fill = "white")+
  facet_wrap(~sex)
  

```



box plot of male and female data set

```{r boxplot}

# boxplot

admission_data %>% 
  ggplot(aes(x=sex,
             y=Sugar_reading))+
  geom_boxplot()+
  geom_jitter(alpha = 0.4)+                   # add data points
  theme(legend.position = "none")             # remove legend


```

```{r calculating-age}
#admission_data$dob = as.Date(admission_data$dob)
admission_data$age = round(age_calc(admission_data$dob, enddate = Sys.Date(), units = 'years' ))
glimpse(admission_data)

```



```{r - scatterplot}


plot(admission_data$age, admission_data$Sugar_reading,
     pch = 19)
 lines(lowess(admission_data$age, admission_data$Sugar_reading ), col = "blue")



```



```{r male-female-relation}
#Scatter plot

plot(admission_data$age, admission_data$Sugar_reading,
     pch = 19,
     col = factor(admission_data$sex))

#Legends
legend("topleft",
       legend = levels(factor(admission_data$sex)),
       pch = 19,
       col = factor(levels(factor(admission_data$sex))))


lines(lowess(admission_data$age, admission_data$Sugar_reading ), col = "blue")

```

```{r}

admission_data %>%
  ggplot(aes(x=age, 
             y=Sugar_reading))+
  geom_point()+
  geom_smooth(se = FALSE)


```




dividing the data set into male and female data set


```{r change-colour}


admission_data %>%
  ggplot(aes(x=age, 
             y=Sugar_reading,
             color =sex))+
  geom_point(alpha=2, size=2)+                   
  geom_smooth(se = FALSE)

```




```{r diabetes-data}

diabetes <- admission_data %>%
  filter(result == "positive")
glimpse(diabetes)


```


```{r}
diabetes %>%
  ggplot(aes(x=age, 
             y=Sugar_reading,
             color =sex))+
  geom_point(alpha=2, size=2)+                   
  geom_smooth(se = FALSE)
```

```{r new-plot}

p <- diabetes %>%
  ggplot(aes(x=age, 
             y=Sugar_reading,
             color =sex))+
  geom_point()+
  geom_smooth(se = FALSE)

ggplotly(p,tooltip = c("x","y"))

```

```{r color-scheme}
p <- diabetes %>%
  ggplot(aes(x=age, 
             y=Sugar_reading,
             color =sex))+
  geom_point()+
  geom_smooth(se = FALSE)+
   ggtitle("Diabetes-related hospital admissions") +  # add title
  scale_color_viridis(discrete=TRUE)         # use the viridis colour scheme
ggplotly(p,tooltip = c("x","y"))

```

this is the end of report